---
title: "Ostend — Coastal Inundation Exposure (with IPCC AR6 scenarios up to 2100)"
format:
  html:
    theme: litera
server: shiny
---

```{r setup, include=FALSE}
# Shiny and UI
library(shiny)
# Spatial vector and raster handling
library(sf)
library(terra)
# Data wrangling
library(dplyr)
library(tidyr)
library(stringr)
# Plotting
library(ggplot2)
# Interactive map
library(leaflet)
# Fast raster-polygon extraction
library(exactextractr)
library(tibble)
library(purrr)
# Axis label formatting
library(scales)
library(leafem)

# Disable spherical geometry to avoid edge-case errors during masking
sf::sf_use_s2(FALSE)
# Suppress terra progress bars in the UI console
terraOptions(progress = 0)

# -------------------------------------------------------------------
# Paths
# -------------------------------------------------------------------
out_dir   <- "outputs/ostend"
city_slug <- "ostend"

# -------------------------------------------------------------------
# Load preprocessed outputs
# -------------------------------------------------------------------
# Elevation raster (DeltaDTM, ~30 m, land-only — ocean pixels are NA)
deltadtm <- terra::rast(file.path(out_dir, "rasters", paste0(city_slug, "_deltadtm.tif")))
# WorldPop 2020 population count raster (used as scaling baseline)
worldpop <- terra::rast(file.path(out_dir, "rasters", paste0(city_slug, "_worldpop.tif")))

# Municipal boundary polygon (used for clipping and area computation)
city_sf <- sf::st_read(
  file.path(out_dir, "vectors", paste0(city_slug, "_boundary.gpkg")),
  quiet = TRUE
) |>
  sf::st_transform(4326) |>
  sf::st_make_valid()

# SLR projections table: columns scenario / quantile / year / slr_m
slr_all <- readRDS(file.path(out_dir, "slr_projections.rds"))


# ESL projections table (SLR + 100-year storm surge per scenario/year/quantile)
esl_all <- readRDS(file.path(out_dir, "esl_projections.rds"))

# The surge value is stationary per city — same for all rows.
# Extract it once as a scalar so the reactive can just add it on top of slr_m.
surge_m_value <- unique(esl_all$surge_m)[1]
gauge_used    <- unique(esl_all$gauge_used)[1]
return_period <- unique(esl_all$return_period)[1]
# -------------------------------------------------------------------
# SSP population projections
# pop_proj_df stores either WorldPop SSP rasters totals or IIASA ratios
# (see preprocessing script). The Shiny uses it to scale the 2020
# WorldPop raster so that exposed-population estimates evolve with
# the selected scenario and year.
# -------------------------------------------------------------------
pop_proj_path <- file.path(out_dir, "pop_projections.rds")
if (file.exists(pop_proj_path)) {
  pop_proj_df <- readRDS(pop_proj_path)
} else {
  # Fallback: hardcoded IIASA SSP ratios for Belgium.
  # This allows the Shiny to run even if preprocessing has not yet
  # produced pop_projections.rds.
  ssp_ratios <- tibble::tribble(
    ~year, ~ssp126, ~ssp245, ~ssp585,
    2020,  1.000,   1.000,   1.000,
    2030,  1.038,   1.042,   1.048,
    2040,  1.065,   1.075,   1.088,
    2050,  1.079,   1.097,   1.120,
    2060,  1.081,   1.108,   1.148,
    2070,  1.072,   1.110,   1.170,
    2080,  1.053,   1.104,   1.184,
    2090,  1.026,   1.092,   1.192,
    2100,  0.994,   1.074,   1.196
  )
  # Reshape to long format so it matches the structure of pop_projections.rds
  pop_proj_df <- ssp_ratios |>
    tidyr::pivot_longer(c(ssp126, ssp245, ssp585),
                        names_to = "scenario", values_to = "ratio") |>
    dplyr::mutate(source = "Scaling IIASA (fallback)")
}

# Helper: return the SSP population growth ratio for a given scenario and year.
# Linearly interpolates between available years (so any year 2020-2100 works).
get_pop_ratio <- function(scen, yr) {
  sub <- dplyr::filter(pop_proj_df, scenario == scen)
  if ("ratio" %in% names(sub)) {
    # Fallback table: ratio column already present
    approx(sub$year, sub$ratio, xout = yr, rule = 2)$y
  } else {
    # Loaded from pop_projections.rds: compute ratio from population column
    base2020 <- sub$population[sub$year == 2020]
    if (length(base2020) == 0 || base2020 == 0) return(1)
    approx(sub$year, sub$population / base2020, xout = yr, rule = 2)$y
  }
}

# -------------------------------------------------------------------
# Population density raster (2020 baseline, used for scaling)
# -------------------------------------------------------------------
# Load from disk if available; otherwise derive it on the fly
dens_path <- file.path(out_dir, "rasters", paste0(city_slug, "_worldpop_density.tif"))
if (file.exists(dens_path)) {
  worldpop_density <- terra::rast(dens_path)
} else {
  # cellSize() gives the exact geodetic area per cell (accounts for cos-latitude distortion)
  pixel_area_km2   <- terra::cellSize(worldpop, unit = "km")
  worldpop_density <- worldpop / pixel_area_km2
  names(worldpop_density) <- "pop_per_km2"
}

# -------------------------------------------------------------------
# Reproject once to WGS84 and clip to city boundary
# -------------------------------------------------------------------
city_vect_ll <- terra::vect(city_sf)

# Elevation: reprojected and clipped — the NA ocean pixels are preserved
deltadtm_city <- terra::mask(
  terra::crop(terra::project(deltadtm, "EPSG:4326"), city_vect_ll),
  city_vect_ll
)
# worldpop_city and worldpop_density_city are 2020 baselines.
# The Shiny scales them by get_pop_ratio() at runtime — they must be kept.
worldpop_city <- terra::mask(
  terra::crop(terra::project(worldpop, "EPSG:4326"), city_vect_ll),
  city_vect_ll
)
worldpop_density_city <- terra::mask(
  terra::crop(terra::project(worldpop_density, "EPSG:4326"), city_vect_ll),
  city_vect_ll
)

# north arrow
north_html <- htmltools::HTML(
  '<div style="background: white; padding: 6px 8px; border: 2px solid #555;
               border-radius: 4px; font-family: sans-serif; text-align: center;
               box-shadow: 0 1px 3px rgba(0,0,0,0.3);">
     <svg width="20" height="28" viewBox="0 0 20 28">
       <polygon points="10,0 18,24 10,18 2,24" fill="#333"/>
     </svg>
     <div style="font-size: 11px; font-weight: bold; color: #333;">N</div>
   </div>'
)

# -------------------------------------------------------------------
# Scenario labels and colors
# -------------------------------------------------------------------
scenario_labels <- c(
  ssp126 = "SSP1-2.6 — Optimistic",
  ssp245 = "SSP2-4.5 — Intermediate",
  ssp585 = "SSP5-8.5 — Pessimistic"
)
scenario_colors <- c(
  ssp126 = "#2166ac",
  ssp245 = "#f4a582",
  ssp585 = "#d6604d"
)

# -------------------------------------------------------------------
# Available years (slider)
# -------------------------------------------------------------------
# Keep only years within the 2020-2100 window used in the UI slider
years_available <- sort(unique(slr_all$year))
years_shiny     <- years_available[years_available >= 2020 & years_available <= 2100]

# -------------------------------------------------------------------
# Map extent
# -------------------------------------------------------------------
# Add a 2.5 km display buffer around the city so the map is not clipped
# right at the municipal boundary
bbox_ll <- sf::st_bbox(
  sf::st_transform(
    sf::st_buffer(sf::st_transform(city_sf, 32631), 2500),
    4326
  )
)

# -------------------------------------------------------------------
# City area in km²
# -------------------------------------------------------------------
# Used to compute the "% of city exposed" statistic
city_area_km2 <- as.numeric(sf::st_area(sf::st_transform(city_sf, 32631))) / 1e6

# -------------------------------------------------------------------
# Helper functions
# -------------------------------------------------------------------

# Ocean-connectivity flood filter.
# Retains only flood pixels contiguous with open ocean (NA pixels in DeltaDTM).
# Isolated inland depressions (e.g. behind dikes or ridges) are excluded.
# See preprocessing script for the full algorithm documentation.
flood_connected <- function(elev_raster, slr_val, n_border = 5L) {
  flooded_land  <- terra::ifel(!is.na(elev_raster) & elev_raster <= slr_val, 1L, NA)
  elev_ext      <- terra::extend(elev_raster, n_border)
  flood_ext     <- terra::extend(flooded_land, n_border)
  ocean_ext     <- terra::ifel(is.na(elev_ext), 1L, NA)
  # Merge flooded-land and ocean into a single binary surface
  combined      <- terra::cover(flood_ext, ocean_ext)
  # Label connected components (8-direction connectivity)
  patches       <- terra::patches(combined, directions = 8L, zeroAsNA = TRUE)
  # Identify patch IDs that contain at least one ocean pixel
  ocean_patch_r <- terra::ifel(!is.na(ocean_ext), patches, NA)
  ocean_ids     <- unique(na.omit(terra::values(ocean_patch_r, mat = FALSE)))
  if (length(ocean_ids) == 0L) {
    warning("flood_connected: no ocean pixels found; returning empty raster.")
    return(terra::ifel(!is.na(elev_raster), NA_integer_, NA_integer_))
  }
  # Remap ocean-connected patches to 1; isolated patches become NA
  rcl           <- cbind(ocean_ids, rep(1L, length(ocean_ids)))
  connected_ext <- terra::classify(patches, rcl, others = NA)
  connected     <- terra::crop(connected_ext, elev_raster)
  # Remove ocean pixels themselves from the output (land-only result)
  result        <- terra::mask(connected, flooded_land, maskvalue = NA)
  names(result) <- "flooded"
  result
}

# Return a binary flood raster (1 = flooded, NA = not flooded).
# connected = TRUE applies the ocean-connectivity filter; FALSE uses the simple bathtub.
flood_raster_from_slr <- function(slr_val, connected = FALSE) {
  if (connected) {
    flood_connected(deltadtm_city, slr_val)
  } else {
    flooded <- deltadtm_city <= slr_val
    # Preserve NA (ocean) pixels so they are not shown as flooded land
    flooded <- terra::ifel(is.na(deltadtm_city), NA, flooded)
    names(flooded) <- "flooded"
    flooded
  }
}

# Convert the binary flood raster to an sf polygon for Leaflet rendering
flood_polygon_from_raster <- function(flooded_raster) {
  tryCatch({
    flooded_only <- terra::ifel(flooded_raster == 1, 1, NA)
    p <- terra::as.polygons(flooded_only, dissolve = TRUE, na.rm = TRUE) |>
      sf::st_as_sf() |>
      sf::st_make_valid() |>
      sf::st_transform(4326)
    if (nrow(p) == 0) NULL else p
  }, error = function(e) NULL)
}

# Compute the total flooded area in km² from the binary raster.
# Uses per-cell geodetic areas to handle WGS84 distortion.
flooded_area_km2_from_raster <- function(flooded_raster) {
  tryCatch({
    flooded_only   <- terra::ifel(flooded_raster == 1, 1, NA)
    area_r         <- terra::cellSize(flooded_only, unit = "km")
    flooded_area_r <- terra::ifel(is.na(flooded_only), NA, area_r)
    terra::global(flooded_area_r, "sum", na.rm = TRUE)[1, 1]
  }, error = function(e) 0)
}

# Population exposure sum.
# pop_raster_scaled: optional SpatRaster already multiplied by the SSP ratio.
# Falls back to the 2020 worldpop_city raster when not supplied.
exposed_population_from_polygon <- function(flood_poly, pop_raster_scaled = NULL) {
  if (is.null(flood_poly) || nrow(flood_poly) == 0) return(0)
  raster_to_use <- if (!is.null(pop_raster_scaled)) pop_raster_scaled else worldpop_city
  tryCatch({
    exactextractr::exact_extract(raster_to_use, flood_poly, "sum") |>
      sum(na.rm = TRUE)
  }, error = function(e) 0)
}
```

## Context

This application maps the potential coastal inundation exposure of Ostend (Belgium) under three IPCC AR6 greenhouse gas emission scenarios and across the 21st century. A static bathtub model is used : a pixel is considered exposed when its elevation (DeltaDTM, 30 m resolution) is at or below the projected sea-level rise relative to the 1995–2014 baseline. The exposed population is estimated from WorldPop data scaled to the selected SSP scenario and year using IIASA SSP demographic projections.

> This is not a hydrodynamic flood model. Storm surge propagation, tides and drainage infrastructure are not accounted for. Results represent a topographic exposure proxy. Two flood modes are available: the **Simple bathtub** marks every pixel below the SLR threshold; **Ocean-connected only** shows only areas connected to the sea and below SLR threshold. Therefore, it does not consider areas protected by dikes to be at risk. **100 year storm surge** adds a stationary surge height (~224 cm at Oostende) on top of the mean SLR. This shows what could be flooded if a 1% probability storm hits in the selected year and scenario. The probability is annual and does not depend on the year chosen.

# How to use the controls

**Projection year** : slide to explore how exposure evolves between 2020 and 2100.

**Emissions scenario** : choose between an optimistic (SSP1-2.6), intermediate (SSP2-4.5), or pessimistic (SSP5-8.5) future.

**Confidence level** : the IPCC projections are probabilistic. The median (50%) is the central estimate. Choosing 83% or 95% means that only 17% or 5% of model runs project more sea-level rise than shown.

## Controls

```{r ui_controls}
fluidRow(
  column(4,
    sliderInput("year", label = "Projection year:",
                min = min(years_shiny), max = max(years_shiny),
                value = 2050, step = 10, sep = "",
                animate = animationOptions(interval = 1200, loop = FALSE))
  ),
  column(4,
    selectInput("scenario", label = "Emissions scenario:",
                choices  = setNames(names(scenario_labels), scenario_labels),
                selected = "ssp245")
  ),
  column(4,
    selectInput("quantile", label = "Confidence level:",
                choices = c("Median (50%)"           = "0.50",
                            "Optimistic (17%)"       = "0.17",
                            "Pessimistic (83%)"      = "0.83",
                            "Very pessimistic (95%)" = "0.95"),
                selected = "0.50")
  )
)
fluidRow(
  column(12,
    div(
      style = "margin-top:6px; padding:8px 12px; background:#f0f4f8;
               border-left:4px solid #2196F3; border-radius:4px;",
      checkboxInput(
        "connectivity",
        label = HTML("<b>Ocean-connected only</b>"),
        value = TRUE
      )
    )
  )
)
fluidRow(
  column(12,
    div(
      style = "margin-top:6px; padding:8px 12px; background:#fff3e0;
               border-left:4px solid #ff9800; border-radius:4px;",
      checkboxInput(
        "include_surge",
        label = HTML(paste0(
          "<b>Include ", return_period,
          "-year storm surge</b> "
        )),
        value = FALSE
      )
    )
  )
)
```

## Interactive map

```{r ui_map}
leafletOutput("flood_map", height = "550px")
```

## Summary statistics

```{r ui_stats}
fluidRow(
  column(3, wellPanel(h4("Sea-level rise"),    textOutput("slr_value"),
                      style = "text-align:center; background:#e8f4f8;")),
  column(3, wellPanel(h4("Flooded area"),      textOutput("flooded_area"),
                      style = "text-align:center; background:#fde0dc;")),
  column(3, wellPanel(h4("Exposed population"), textOutput("pop_exposed"),
                      style = "text-align:center; background:#fde0dc;")),
  column(3, wellPanel(h4("Exposed city area"), textOutput("pct_exposed"),
                      style = "text-align:center; background:#fde0dc;"))
)
```

## Temporal evolution

```{r ui_timeseries}
plotOutput("ts_plot", height = "320px")
```

## Shiny server

```{r server}
#| context: server
# -------------------------------------------------------------------
# Reactive: total water-level value for selected scenario/year/quantile.
# When `include_surge` is ON, the stationary storm-surge return level is
# added on top of the mean SLR. The water-level returned by this reactive
# is what drives the flood map (bathtub threshold) and the statistics.
# -------------------------------------------------------------------
slr_reactive <- reactive({
  base_slr <- slr_all |>
    dplyr::filter(
      scenario == input$scenario,
      year     == input$year,
      abs(quantile - as.numeric(input$quantile)) < 0.001
    ) |>
    dplyr::pull(slr_m) |>
    dplyr::first()

  if (isTRUE(input$include_surge)) {
    base_slr + surge_m_value
  } else {
    base_slr
  }
})

# -------------------------------------------------------------------
# Reactive: flood products + SSP-scaled population rasters
# -------------------------------------------------------------------
flood_reactive <- reactive({
  slr_val <- slr_reactive()
  # Abort early if SLR value is unavailable (e.g. missing scenario/year combination)
  if (is.null(slr_val) || length(slr_val) == 0 || is.na(slr_val)) return(NULL)

  # 1. Flood raster (simple bathtub or ocean-connected)
  use_connectivity <- isTRUE(input$connectivity)
  flooded    <- flood_raster_from_slr(slr_val, connected = use_connectivity)
  flood_poly <- flood_polygon_from_raster(flooded)

  # 2. SSP population ratio for the selected scenario and year.
  #    This scales the 2020 WorldPop baseline so that population exposure
  #    reflects the demographic trajectory of the chosen scenario.
  pop_ratio <- tryCatch(
    get_pop_ratio(input$scenario, as.numeric(input$year)),
    error = function(e) 1  # default to ratio = 1 (no change) on error
  )

  # 3. Scaled rasters:
  #    - worldpop_density_proj: density (inhab/km²) for the map gradient
  #    - worldpop_count_proj:   count (persons) for the exposure statistic
  worldpop_density_proj <- worldpop_density_city * pop_ratio
  worldpop_count_proj   <- worldpop_city         * pop_ratio

  # 4. Mask density raster to the flooded zone for the population gradient layer
  pop_flooded <- if (!is.null(flood_poly) && nrow(flood_poly) > 0) {
    tryCatch(
      terra::mask(worldpop_density_proj, terra::vect(flood_poly)),
      error = function(e) NULL
    )
  } else NULL

  list(
    raster           = flooded,
    poly             = flood_poly,
    slr_m            = slr_val,
    pop_ratio        = pop_ratio,
    pop_flooded      = pop_flooded,         # density masked to flood zone (map)
    pop_count_scaled = worldpop_count_proj  # full count raster scaled to SSP (stats)
  )
})

# -------------------------------------------------------------------
# Leaflet map (static base — flood layers updated via leafletProxy)
# -------------------------------------------------------------------
output$flood_map <- renderLeaflet({
  leaflet() |>
    addProviderTiles("Esri.WorldImagery",
                     options = providerTileOptions(maxZoom = 18)) |>
    addProviderTiles("CartoDB.PositronOnlyLabels") |>
    fitBounds(bbox_ll[["xmin"]], bbox_ll[["ymin"]],
              bbox_ll[["xmax"]], bbox_ll[["ymax"]]) |>
    addPolygons(data = city_sf, fill = FALSE,
                color = "#333333", weight = 2.5,
                group = "City boundary") |>
    addScaleBar(position = "bottomleft",
                options = scaleBarOptions(metric = TRUE,
                                          imperial = FALSE,
                                          maxWidth = 150)) |>
    addControl(html = north_html, position = "topright") |>
    addLayersControl(
      overlayGroups = c("City boundary", "Flooded area", "Exposed population"),
      options = layersControlOptions(collapsed = FALSE)
    )
})

# -------------------------------------------------------------------
# Update flood + population layers without rebuilding the map
# -------------------------------------------------------------------
observe({
  req(input$scenario, input$year, input$quantile)
  res <- flood_reactive()

  # Clear previous flood and population layers before redrawing
  leafletProxy("flood_map") |>
    clearGroup("Flooded area") |>
    clearGroup("Exposed population") |>
    removeControl("pop_legend")

  if (!is.null(res) && !is.null(res$poly) && nrow(res$poly) > 0) {

    # Flood extent polygon — semi-transparent blue fill
    leafletProxy("flood_map") |>
      addPolygons(
        data        = res$poly,
        fillColor   = "#2196F3",
        fillOpacity = 0.25,
        color       = "#0d47a1",
        weight      = 0.8,
        group       = "Flooded area",
        popup       = paste0(
          "<b>Potentially flooded area</b><br>",
          "Sea-level rise: +", round(res$slr_m * 100, 1), " cm<br>",
          "Scenario: ", scenario_labels[input$scenario], "<br>",
          "Year: ", input$year, "<br>",
          "Mode: ", if (isTRUE(input$connectivity)) "Ocean-connected only"
                    else "Simple bathtub", "<br>",
          "Population ratio vs 2020: x", round(res$pop_ratio, 3)
        )
      )

    # Population density gradient overlay
    if (!is.null(res$pop_flooded)) {
      pop_vals     <- terra::values(res$pop_flooded, na.rm = TRUE)
      pop_vals_pos <- pop_vals[pop_vals > 0]
      # Separate zero-population pixels (grey) from populated pixels (gradient)
      pop_zero     <- terra::ifel(res$pop_flooded == 0, 1L, NA)
      pop_has_pop  <- terra::ifel(res$pop_flooded >  0, res$pop_flooded, NA)
      has_zero     <- sum(!is.na(terra::values(pop_zero))) > 0
      has_pop      <- length(pop_vals_pos) > 0 && max(pop_vals_pos, na.rm = TRUE) > 0

      lproxy <- leafletProxy("flood_map")

      # Grey layer: flooded pixels with no population
      if (has_zero) {
        lproxy <- lproxy |>
          addRasterImage(pop_zero, colors = "#BBBBBB",
                         opacity = 0.65, group = "Exposed population")
      }

      # Yellow-red gradient: flooded pixels with population
      if (has_pop) {
        # Domain is restricted to positive values so zero pixels don't pull the scale
        pal_pop <- colorNumeric(
          palette  = c("#FFFF33", "#FF8C00", "#CC0000"),
          domain   = pop_vals_pos,
          na.color = "transparent"
        )
        lproxy <- lproxy |>
          addRasterImage(pop_has_pop, colors = pal_pop,
                         opacity = 0.85, group = "Exposed population")
      }

      # Build legend: grey entry first (if any), then three gradient classes
      legend_colors <- c(
        if (has_zero) "#BBBBBB",
        if (has_pop)  c("#FFFF33", "#FF8C00", "#CC0000")
      )
      legend_labels <- c(
        if (has_zero) "No population",
        if (has_pop)  c(
          paste0("Low (< ",    round(quantile(pop_vals_pos, 0.33, na.rm = TRUE)), ")"),
          paste0("Medium (< ", round(quantile(pop_vals_pos, 0.66, na.rm = TRUE)), ")"),
          paste0("High (>=",   round(quantile(pop_vals_pos, 0.66, na.rm = TRUE)), ")")
        )
      )
      lproxy |>
        addLegend(position = "bottomright",
                  colors = legend_colors, labels = legend_labels,
                  title   = paste0("Population density<br>(inhab/km²)<br>",
                                   input$scenario, " — ", input$year),
                  layerId = "pop_legend", opacity = 0.85)
    }
  }
})

# -------------------------------------------------------------------
# Summary statistics
# -------------------------------------------------------------------
output$slr_value <- renderText({
  slr_val <- slr_reactive()
  if (is.null(slr_val) || is.na(slr_val)) return("N/A")
  if (isTRUE(input$include_surge)) {
    paste0("+", round(slr_val * 100, 1), " cm",
           " (SLR + ", return_period, "-yr surge)")
  } else {
    paste0("+", round(slr_val * 100, 1), " cm (SLR only)")
  }
})

output$flooded_area <- renderText({
  res <- flood_reactive()
  if (is.null(res)) return("0 km2")
  area <- flooded_area_km2_from_raster(res$raster)
  paste0(round(area, 2), " km2")
})

output$pop_exposed <- renderText({
  res <- flood_reactive()
  if (is.null(res)) return("0 people")
  # Uses the SSP-scaled count raster so the number reflects the chosen scenario+year
  pop <- exposed_population_from_polygon(res$poly, res$pop_count_scaled)
  paste0(format(round(pop), big.mark = ","), " people")
})

output$pct_exposed <- renderText({
  res <- flood_reactive()
  if (is.null(res)) return("0 %")
  area <- flooded_area_km2_from_raster(res$raster)
  paste0(round(area / city_area_km2 * 100, 1), " %")
})

# -------------------------------------------------------------------
# Time-series plot (SLR trajectories with selected point highlighted)
# -------------------------------------------------------------------
output$ts_plot <- renderPlot({
  q_val <- as.numeric(input$quantile)

  # Prepare data: keep 5th, 50th, 95th percentiles over 2020-2100
  slr_plot_df <- slr_all |>
    dplyr::filter(quantile %in% c(0.05, 0.50, 0.95),
                  year >= 2020, year <= 2100) |>
    dplyr::mutate(scenario_label = scenario_labels[scenario]) |>
    dplyr::arrange(scenario, quantile, year)

  # Wide format needed for geom_ribbon (ymin = 5th, ymax = 95th)
  ribbon_df <- slr_plot_df |>
    dplyr::filter(quantile %in% c(0.05, 0.95)) |>
    dplyr::select(-scenario_label) |>
    tidyr::pivot_wider(names_from = quantile, values_from = slr_m) |>
    dplyr::arrange(scenario, year)

  # Single point for the currently selected scenario / year / quantile
  selected_pt <- slr_all |>
    dplyr::filter(scenario == input$scenario,
                  year     == input$year,
                  abs(quantile - q_val) < 0.001)

  ggplot(slr_plot_df |> dplyr::filter(quantile == 0.50),
         aes(x = year, y = slr_m * 100, color = scenario)) +
    # Uncertainty ribbon: 5th to 95th percentile
    geom_ribbon(data = ribbon_df,
                aes(x = year, ymin = `0.05` * 100, ymax = `0.95` * 100, fill = scenario),
                color = NA, alpha = 0.15, inherit.aes = FALSE) +
    # Median trajectory line
    geom_line(linewidth = 1) +
    # Highlighted point for the user-selected combination
    geom_point(data = selected_pt,
               aes(x = year, y = slr_m * 100),
               size = 5, shape = 21, fill = "white",
               color = scenario_colors[input$scenario],
               stroke = 2, inherit.aes = FALSE) +
    scale_color_manual(values = scenario_colors, labels = scenario_labels) +
    scale_fill_manual( values = scenario_colors, labels = scenario_labels) +
    scale_x_continuous(breaks = seq(2020, 2100, by = 10)) +
    scale_y_continuous(labels = scales::label_number(suffix = " cm")) +
    labs(title    = "IPCC AR6 sea-level rise projections for Ostend",
         subtitle = "Line = median (50%) | Ribbon = 5th to 95th percentile",
         x = "Year", y = "Sea-level rise (cm)",
         color = "Scenario", fill = "Scenario",
         caption = "Source: IPCC AR6 | Zenodo 6382554") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())
})
```
